Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-03-14 lics2021-rebuttal-51.txt

All three reviewers see the paper in a positive light, and we thank
them for their feedback. The complaints of reviewer 1 concern only the
presentation, which can easily improved with all the helpful comments
of reviewer 1.

Below we give more detailed answers to the questions raised by
reviewers 1 and 2.

*********************************************************************
Answer to review 1:
===================

We are very grateful for the detailed comments, and we will happily
follow all suggestions. In particular, we will include an outline of
the paper, give more intuitive explanations and a broader overview of
the proof. Also the remark about equality is helpful.

> Question p.3: require that adjacent vertices to have different colors?

Indeed, in [18] there is this requirement, but it is not needed
because of the correctness criterion. Therefore we did not include it.

> Question p.3: can different vertices have the same label?

Yes.

> Question p.3: What about the atoms True and False in the fographs?

Indeed, in classical logic they are not needed for semantic
soundness and completeness, because

True = a v -a  and  False = a ^ -a

However, they are part of the language, and they also make the proof
systems simpler to use. Therefore we decided to include them.

> Question p.7: is \overline{S{A}} the complement of S{A} ?

Yes.

> Question p.10: A\cdot is variable renaming?

Yes.

> Question p.10: why "resource management"?

We see a combinatorial proof as a way to separate the linear part of a
proof (in the sense of linear logic) from the resource management of a
proof (in the sense that contraction and weakening duplicate and
delete resources). As this section is about contraction and weakening,
we chose this title. We agree that this needs to be better explained.

> Question p.11: "strong" homomorphism?

No, we do not have this condition. Note that this lemma is only about
weakening, and therefore the needed conditions are full and
injective. The next lemma is about contraction, where we need
surjectivity.

*********************************************************************
Answer to review 2:
===================

Question about the proof of Lemma 30:

Indeed, the structural equivalences concerning the quantifiers are not
discussed in [35]. But in our case, the equivalences concerning the
quantifiers are admissible in the system without the exists-rule. In
particular, in the example given in the review, a proof of

(forall x. A) \/ B

can be straightforwardly transformed into a proof of

forall x. (A \/ B)

(if x is vacuous in B)

We agree that this should be explained better.

### 2021-03-15 lics2021-rebuttal-51.txt

All three reviewers see the paper in a positive light, and we thank
them for their feedback. The complaints of reviewer 1 concern only the
presentation, which can easily improved with all the helpful comments
of reviewer 1.

Below we give more detailed answers to the questions raised by
reviewers 1 and 2.

*********************************************************************
Answer to review 1:
===================

We are very grateful for the detailed comments, and we will happily
follow all suggestions. We will include an outline of the paper, give
more intuitive explanations, and provide a broad overview of
the proof. The remark about equality is helpful.

@@

> Question p.3: require that adjacent vertices to have different colors?

It was included in [18] because it is part of the standard definition of
colouring in graph theory. However, it is redundant, hence optional,
since it is implied by the preclusion of bimatchings: if vw is edge
between like-coloured vertices (v~w), then {v,w} is a bimatching.

@@

> Question p.10: why "resource management"?

We see a combinatorial proof as a way to separate the linear part of a
proof (in the sense of linear logic) from the resource management of a
proof (in the sense that contraction and weakening duplicate and
delete resources). As this section is about contraction and weakening,
we chose this title. We agree that this needs to be better explained.

> Question p.11: "strong" homomorphism?

No, we do not have this condition. This lemma is only about
weakening, so the requisite conditions are full and injective. The next
lemma is about contraction, where we need surjectivity.

### 2021-03-15 lics2021-rebuttal-51.txt

All three reviewers see the paper in a positive light, and we thank
them for their feedback. The complaints of reviewer 1 concern only the
presentation, which can easily improved with all the helpful comments
of reviewer 1.

Below we give more detailed answers to the questions raised by
reviewers 1 and 2.

*********************************************************************
Answer to review 1:
===================

We are very grateful for the detailed comments, and we will happily
follow all suggestions. In particular, we will include an outline of
the paper, give more intuitive explanations and a broader overview of
the main proof. Also the remark about equality is helpful.

### 2021-03-15 lics2021-rebuttal-51.txt

All three reviewers see the paper in a positive light, and we thank
them for their feedback. The complaints of reviewer 1 concern only the
presentation, which can easily improved with all the helpful comments
of reviewer 1.

Below we give more detailed answers to the questions raised by
reviewers 1 and 2.

*********************************************************************
Answer to review 1:
===================

We are very grateful for the detailed comments, and we will happily
follow all suggestions. We will include an outline of the paper, give
more intuitive explanations, and provide a broad overview of
the main proof. The remark about equality is helpful.

### 2021-04-21 LICS2021-final.tex
starting the final version
tiny modifications

%%line and pagenumbers to be removed for final version
\usepackage[switch]{lineno}
\pagestyle{plain}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{document}
\vlnosmallleftlabels%
\bstctlcite{refs:BSTcontrol}%
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Combinatorial Proofs and Decomposition Theorems for First-order Logic\vspace{-3ex}}

@@

%% \author{\IEEEauthorblockN{Dominic J. D. Hughes}
%% \IEEEauthorblockA{}
%% \and
%% \IEEEauthorblockN{Lutz Straßburger}
%% \IEEEauthorblockA{Inria, Equipe Partout\\
%% Ecole Polytechnique, LIX \\ %UMR 7161\\
%% France}
%% \and
%% \IEEEauthorblockN{Jui-Hsuan Wu}
%% \IEEEauthorblockA{Ecole Normale Supérieure\\
%% France}}

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\linenumbers

@@

Fix pairwise disjoint countably infinite sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability we often omit parentheses and write simply $ft_1 \dots t_n$ or $pt_1 \dots t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

@@

\begin{proof}
   By \cite[Proposition~7.6.1]{str:07:RTA}, we have a derivation
   $\vlder{\set{\wrD,\PE\fequ}}{\DDeri}{\PE H}{\PE G}$. In order to
   lift $\DDeri$, we need to reorganize the instances of
   $\wD$. If $H$ contains a subformula $\forall x.A$ which is not
   present in $G$, the $\wD$-instances in $\DDeri$ could introduce the
   parts of the propositional encoding $x\vlor A$ independently.  We
   say that an instance $\rr_1$ of $\wD$ in $\Deri$ is \emph{in the
   scope} of an instance $\rr_2$ of $\wD$ if $\rr_1$ introduced
   formulas that contain a free variable $x$ (i.e., $x$ occurs in a
   term in a predicate) and $\rr_2$ introduces the atom $x$ as a
   subformula (i.e. the propositional encoding of the binder $x$).  We
   can now permute the $\wD$-instances in $\DDeri$ such that whenever
   a rule instance $\rr_1$ is in the scope on an instance $\rr_2$, then
   $\rr_2$ occurs below $\rr_1$ in $\DDeri$.
   Then we can lift $\DDeri$ stepwise. First, observe that each line of
   $\DDeri$ is $\PE\fequ$-equivalent to the propositional encoding $\PE P$ of a
   first-order formula $P$.  We now have to show
   that each instance of $\wrD$ in $\DDeri$ is indeed the image of a correct
   application of $\wD$ or $\wfaD$ in first-order logic.
   If we have a $\wD$ of the form
   \begin{equation*}
     \scalebox{1}{$
     \vlinf{\wD}{}{\PE S\cons{x\vlor \PE A}}{\PE S\cons{\PE A}}
     \qquor
     \vlinf{\wD}{}{\PE S\cons{(x\vlor \PE B)\vlor \PE A}}{\PE S\cons{\PE A}}
     $}
   \end{equation*}
   then $x$ cannot occur freely in $A$, as otherwise the fibration
   property would be violated. We can therefore lift these instances to
   \begin{equation*}
      \scalebox{1}{$
    \vlinf{\wfaD}{}{S\cons{\forall x. A}}{S\cons{A}}
     \qquor
     \vlinf{\wD}{}{S\cons{(\forall x. B)\vlor A}}{S\cons{A}}
     $}
   \end{equation*}
   respectively.     
  %% A subformula of $\PE B$ is called \emph{weak} if it has been
  %% introduced by a weakening. All subformulas of a weak formula are
  %% also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  %% \emph{weak pair} if $\PE B\fequ\Scons{B'\vlor B''}$ for some context
  %% $\Sconhole$. We can assume without loss of generality that
  %% whenever weak subformulas $B'$ and $B''$ form a weak pair, they have
  %% been introduced by the same instance of $\wrD$ in
  %% $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  %% brought into this form by simple rule permutations, and then
  %% collapsing several weakenings into a single one.}  Now we show that
  %% $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  %% weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  %% delete all atoms in the scope of the corresponding quantifier,
  %% because $\phi'$ is a fibration on the binding graph. Hence, each line
  If a  weakening happens inside a subformula $x\vlor \PE C$ or $x\vlan
  \PE C$ in $\DDeri$, then there are the following cases:
  \begin{equation*}%\small
    \scalebox{.9}{$
    \vlinf{\wD}{}{\PE S\cons{x\vlor\PE  D\vlor \PE C}}{\PE S\cons{x\vlor\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{x\vlan (\PE D\vlor \PE C)}}{\PE S\cons{x\vlan\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{(x\vlor\PE  D)\vlan \PE C}}{\PE S\cons{x\vlan \PE C}}
    $}
  \end{equation*}
  The first two cases can be lifted to
  \begin{equation*}
    \vlinf{\wD}{}{S\cons{\forall x.( D\vlor C)}}{S\cons{\forall x.  C}}
    \quand
     \vlinf{\wD}{}{S\cons{\exists x.( D\vlor C)}}{S\cons{\exists x.  C}}
  \end{equation*}
  respectively. But in the third case,  an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur. All other situations can be lifted trivially, giving us
  $\vlder{\set{\wrD,\wfaD,\fequ}}{\Deri}{H}{G}$ as desired.
\end{proof}

>>>>

%%line and pagenumbers to be removed for final version
%\usepackage[switch]{lineno}
\pagestyle{plain}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{document}
\vlnosmallleftlabels%
\bstctlcite{refs:BSTcontrol}%
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Combinatorial Proofs and Decomposition Theorems for First-order Logic}

@@

\author{\IEEEauthorblockN{Dominic J. D. Hughes}
\IEEEauthorblockA{}
\and
\IEEEauthorblockN{Lutz Straßburger}
\IEEEauthorblockA{Inria, Equipe Partout\\
Ecole Polytechnique, LIX \\ %UMR 7161\\
France}
\and
\IEEEauthorblockN{Jui-Hsuan Wu}
\IEEEauthorblockA{Ecole Normale Supérieure\\
France}}

@@

% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
%
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
%\IEEEpeerreviewmaketitle

%\linenumbers

@@

Fix pairwise disjoint countably infinite sets $\VAR=\set{x,y,z,\ldots}$ of variables,
$\FUN=\set{f,g,\ldots}$ of function symbols, and 
$\PRED=\set{p,q,\ldots}$ of predicate symbols. Each function symbol and
each predicate symbol has a finite arity. Each predicate symbol $p$ has a
\bfit{dual}
$\dual{p}$. The grammars below generate
the set $\TERM$ of \bfit{terms}, denoted by $s,t,u,\ldots$, the set
$\ATOM$ of \bfit{atoms}, denoted by $a,b,c,\ldots$, and the set
$\FORM$ of \bfit{formulas}, denoted by $A,B,C,\ldots$:
\begin{equation*}
  \begin{array}{r@{~}l}
    t &\coloneqq~ x \mid f(t_1,\dots,t_n)
    \\[.5ex]
    a &\coloneqq~ \ttt\mid\fff\mid p(t_1,\dots,t_n) \mid \dual p(t_1,\dots,t_n)
    \\[.5ex]
    A &\coloneqq~ a \mid A\vlan A \mid A\vlor A \mid \exists x.A \mid \forall x.A
\end{array}
\end{equation*}
where the arity of $f$ and $p$ is $n$.
For better readability we often omit parentheses and write simply $ft_1 \dots t_n$ or $pt_1 \dots t_n$.
We consider the truth constants
$\ttt$ (\emph{true}) and $\fff$
(\emph{false}) as additional atoms, and we consider all formulas in negation
normal form, where \bfit{negation} $(\dual\cdot)$ is defined on
atoms and formulas via De Morgan's laws:
\begin{equation*}
  \begin{array}{ccc}
    \dual{\dual a} = a
    &
  \dual\ttt =\fff &
  \dual{p(t_1,\ldots,t_n)} = \dual{p}(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual\fff = \ttt &
  \dual{\dual p(t_1,\ldots,t_n)} = p(t_1,\ldots,t_n) \\[.5ex]
  &
  \dual{\exists x.A} = \forall x.\dual{A}
  &
  \dual{A \vlan B} = \dual{A} \ \vlor \ \dual{B} 
  \\[.5ex]
  &
  \dual{\forall x.A} = \exists x.\dual{A}
  &
%  \\[.5ex]
  \dual{A \vlor B} = \dual{A} \ \vlan \ \dual{B}
  \end{array}
\end{equation*}
Then we write $A\implies B$ as abbreviation for $\dual A\vlor B$.

@@

\begin{proof}
   By \cite[Proposition~7.6.1]{str:07:RTA}, we have a derivation
   $\vlder{\set{\wrD,\PE\fequ}}{\DDeri}{\PE H}{\PE G}$. In order to
   lift $\DDeri$, we need to reorganize the instances of
   $\wD$. If $H$ contains a subformula $\forall x.A$ which is not
   present in $G$, the $\wD$-instances in $\DDeri$ could introduce the
   parts of the propositional encoding $x\vlor A$ independently.  We
   say that an instance $\rr_1$ of $\wD$ in $\Deri$ is \emph{in the
   scope} of an instance $\rr_2$ of $\wD$ if $\rr_1$ introduced
   formulas that contain a free variable $x$ (i.e., $x$ occurs in a
   term in a predicate) and $\rr_2$ introduces the atom $x$ as a
   subformula (i.e. the propositional encoding of the binder $x$).  We
   can now permute the $\wD$-instances in $\DDeri$ such that whenever
   a rule instance $\rr_1$ is in the scope of an instance $\rr_2$, then
   $\rr_2$ occurs below $\rr_1$ in $\DDeri$.
   Then we can lift $\DDeri$ stepwise. First, observe that each line of
   $\DDeri$ is $\PE\fequ$-equivalent to the propositional encoding $\PE P$ of a
   first-order formula $P$.  We now have to show
   that each instance of $\wrD$ in $\DDeri$ is indeed the image of a correct
   application of $\wD$ or $\wfaD$ in first-order logic.
   If we have a $\wD$ of the form
   \begin{equation*}
     \scalebox{1}{$
     \vlinf{\wD}{}{\PE S\cons{x\vlor \PE A}}{\PE S\cons{\PE A}}
     \qquor
     \vlinf{\wD}{}{\PE S\cons{(x\vlor \PE B)\vlor \PE A}}{\PE S\cons{\PE A}}
     $}
   \end{equation*}
   then $x$ cannot occur freely in $A$, as otherwise the fibration
   property would be violated. We can therefore lift these instances to
   \begin{equation*}
      \scalebox{1}{$
    \vlinf{\wfaD}{}{S\cons{\forall x. A}}{S\cons{A}}
     \qquor
     \vlinf{\wD}{}{S\cons{(\forall x. B)\vlor A}}{S\cons{A}}
     $}
   \end{equation*}
   respectively.     
  %% A subformula of $\PE B$ is called \emph{weak} if it has been
  %% introduced by a weakening. All subformulas of a weak formula are
  %% also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  %% \emph{weak pair} if $\PE B\fequ\Scons{B'\vlor B''}$ for some context
  %% $\Sconhole$. We can assume without loss of generality that
  %% whenever weak subformulas $B'$ and $B''$ form a weak pair, they have
  %% been introduced by the same instance of $\wrD$ in
  %% $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  %% brought into this form by simple rule permutations, and then
  %% collapsing several weakenings into a single one.}  Now we show that
  %% $\PE\Deri_1$ can be lifted. For this, observe that whenever a
  %% weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it must also
  %% delete all atoms in the scope of the corresponding quantifier,
  %% because $\phi'$ is a fibration on the binding graph. Hence, each line
  If a  weakening happens inside a subformula $x\vlor \PE C$ or $x\vlan
  \PE C$ in $\DDeri$, then there are the following cases:
  \begin{equation*}%\small
    \scalebox{.9}{$
    \vlinf{\wD}{}{\PE S\cons{x\vlor\PE  D\vlor \PE C}}{\PE S\cons{x\vlor\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{x\vlan (\PE D\vlor \PE C)}}{\PE S\cons{x\vlan\PE  C}}
    \quad
    \vlinf{\wD}{}{\PE S\cons{(x\vlor\PE  D)\vlan \PE C}}{\PE S\cons{x\vlan \PE C}}
    $}
  \end{equation*}
  The first two cases can be lifted to
  \begin{equation*}
    \vlinf{\wD}{}{S\cons{\forall x.( D\vlor C)}}{S\cons{\forall x.  C}}
    \quand
     \vlinf{\wD}{}{S\cons{\exists x.( D\vlor C)}}{S\cons{\exists x.  C}}
  \end{equation*}
  respectively. But in the third case,  an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur. All other situations can be lifted trivially, giving us
  $\vlder{\set{\wrD,\wfaD,\fequ}}{\Deri}{H}{G}$ as desired.
\end{proof}
